### 2008-12-19 talk/ecoop2009/rainbow.tex
eh. not that many references were left to the rainbow interp anyway. die die
make a figure
add tlc example about virtuals

It is not always possible to keep structures virtual.  The main
situation in which it needs to be "forced" (i.e. actually allocated at
run-time) is when the pointer escapes to some non-virtual location like
a field of a real heap structure.

Virtual structures still avoid the run-time allocation of most
short-lived objects, even in non-trivial situations.  The following
example shows a typical case.  The TLC interpreter implements application-level
integers as boxes – instances of an \texttt{IntObject} class with a single
field.

The PyPy Python interpreter
implements application-level integers as boxes – instances of a
\texttt{W\_IntObject} class with a single \texttt{intval} field.  Here is the
addition of two integers:

XXX needs to use TLC examples
\begin{verbatim}
    def add(w1, w2):            # w1, w2 are W_IntObject instances
        value1 = w1.intval
        value2 = w2.intval
        result = value1 + value2
        return W_IntObject(result)
\end{verbatim}

XXX kill the rest?!‽
When interpreting the bytecode for \texttt{a+b+c}, two calls to \texttt{add()} are
issued; the intermediate \texttt{W\_IntObject} instance is built by the first
call and thrown away after the second call.  By contrast, when the
interpreter is turned into a compiler, the construction of the
\texttt{W\_IntObject} object leads to a virtual structure whose \texttt{intval}
field directly references the register in which the run-time addition
put its result.  This location is read out of the virtual structure at
the beginning of the second \texttt{add()}, and the second run-time addition
directly operates on the same register.

An interesting effect of virtual structures is that they play nicely with
promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
function for integers, it must first determine that the two arguments
are indeed integer objects.  In the corresponding dispatch logic, we
have added two hints to promote the type of each of the two arguments.
This produces a compiler that has the following behavior: in the general
case, the expression \texttt{a+b} will generate two consecutive run-time
switches followed by the residual code of the proper version of
\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
the intermediate value will contain a compile-time constant as type.
Promoting a compile-time constant is trivial – no run-time code is
generated.  The whole expression \texttt{a+b+c} thus only requires three
switches instead of four.  It is easy to see that even more switches can
be skipped in larger examples; typically, in a tight loop manipulating
only integers, all objects are virtual structures for the compiler and
the residual code is theoretically optimal – all type propagation and
boxing/unboxing occurs at compile-time.

@@

\begin{itemize}
  \item (compile time): the rainbow interpreter produces machine code until it
    hits a promotion point; e.g.::

>>>>

It is not always possible to keep structures virtual.  The main
situation in which it needs to be "forced" (i.e. actually allocated at
run-time) is when the pointer escapes to some non-virtual location like
a field of a real heap structure.  Virtual structures still avoid the run-time
 allocation of most short-lived objects, even in non-trivial situations.

In addition to virtual structures, the compiler can also handle virtual
containers, namely lists and dictionaries \footnote{(R)Python's dictionaries
  are equivalent to .NET \lstinline{Hashtable}s}.  If the indexing operations
can be evaluated at compile-time (i.e., if the variables holding the indexes
are green), the compiler internally keeps track of the state of the container
and store the items as local variables.

Look again at figure \ref{fig:tlc-folded}: the list in the \lstinline{stack}
variable never escapes from the function.  Moreover, all the indexing
operations (either done explicitly or implicitly by \lstinline{append} and
\lstinline{pop}) are evaluable at compile-time.  Thus, the list is kept
\emph{virtual} and its elements are stored in variables $v_n$, where $n$
represents the index in the list.  Figure \ref{fig:tlc-folded-virtualized}
show how the resulting code looks like; to ease the reading, the state of the
\lstinline{stack} as kept by the compiler is shown in the comments.

\begin{figure}[h]
\begin{center}
\input{tlc-folded-virtualized.py}
\caption{The result of virtualizing the \lstinline{stack} list}
\label{fig:tlc-main}
\end{center}
\end{figure}

XXX kill the rest?!‽
An interesting effect of virtual structures is that they play nicely with
promotion.  Indeed, before the interpreter can call the proper \texttt{add()}
function for integers, it must first determine that the two arguments
are indeed integer objects.  In the corresponding dispatch logic, we
have added two hints to promote the type of each of the two arguments.
This produces a compiler that has the following behavior: in the general
case, the expression \texttt{a+b} will generate two consecutive run-time
switches followed by the residual code of the proper version of
\texttt{add()}.  However, in \texttt{a+b+c}, the virtual structure representing
the intermediate value will contain a compile-time constant as type.
Promoting a compile-time constant is trivial – no run-time code is
generated.  The whole expression \texttt{a+b+c} thus only requires three
switches instead of four.  It is easy to see that even more switches can
be skipped in larger examples; typically, in a tight loop manipulating
only integers, all objects are virtual structures for the compiler and
the residual code is theoretically optimal – all type propagation and
boxing/unboxing occurs at compile-time.

@@

\begin{itemize}
  %XXX remove mention of rainbow interp. but this needs to be rewritten anyway
  \item (compile time): the rainbow interpreter produces machine code until it
    hits a promotion point; e.g.::
